Nested Stochastic Block Models Applied to the Analysis of Single Cell Data

### 2021-04-27 manuscript.tex
added figure S6
Final retouch

A key step in the analysis of scRNA-seq data and, more in general, of single cell data, is the identification of cell populations, that is groups of cells sharing similar properties. Several approaches have been proposed to achieve this task, based on well established clustering techniques \cite{wang_2017, lin_2017}, consensus clustering \cite{huh_2020, kiselev_2017, Ranjan_2021} and deep learning \cite{li_2020}; many more have been recently reviewed \cite{krzak_2019, kiselev_2019} and benchmarked \cite{du_2018}. As the popularity of single cell analysis frameworks \texttt{Seurat} \cite{butler_2018} and \texttt{scanpy} \cite{wolf_2018} raised, methods based instead on graph partitioning became the \emph{de facto} standards. Such methods require the construction of a cell neighbourhood graph (\emph{e.g.} by \emph{k }Nearest Neighbours, \emph{k}NN, or shared Nearest Neighbors, \emph{s}NN). Encoding cell-to-cell similarities into graphs has practical advantages beyond clustering, as many algorithms for graph analysis can be applied and interpreted in a biological way. A notable example is the analysis of cell trajectories which can be derived from the analysis of Markov processes traversing the NN graph \cite{Setty_2019, Lange_2020}. In another context, computation of RNA moments in scRNA velocity is also based on the NN graph structure \cite{Bergen_2020}. Arguably, the biggest utility of NN structure is the possibility to identify cell groups by partitioning the graph into communities; this is typically achieved using the Louvain method \cite{blondel_2008}, a fast algorithm for optimisation of graph modularity. While fast, this method does not guarantee the identification of internally connected communities. To overcome its limits, a more recent approach, the Leiden algorithm \cite{traag_2019}, has been implemented and it has been quickly adopted in the analysis of single cell data, for example by \texttt{scanpy} \cite{wolf_2018} and \texttt{PhenoGraph} \cite{levine_2015}. In addition to Newman's modularity \cite{newman_2004}, other definitions currently used in single cell analysis make use of a resolution parameter \cite{traag_2011, reichardt_2006}. In lay terms, resolution works as a threshold on the density within communities: lowering the resolution results in less and sparser communities and \emph{vice versa}. Identification of an appropriate resolution has been recognised as a major issue \cite{lhnemann_2020}, also because it requires the definition of a mathematical property (clusters) over biological entities (the cell groups), with little formal description of the latter. In addition, the larger the dataset, the harder is to identify small cell groups, as a consequence of the well-known resolution limit \cite{fortunato_2007}. Moreover, it has been demonstrated that random networks can have modularity \cite{guimer_2004} and its optimisation is incapable of separating actual structure from those arising simply of statistical fluctuations of the null model. Lastly, it is a common error to assume that the resolution parameter reflects a hierarchical structure of the communities in the graph when, in general, this is not rigorously true. Additional solutions to cell group identification from NN graphs have been proposed, introducing resampling techniques \cite{baran_2019, Tang_Sackton_2020} or clique analysis \cite{xu_2015}. It has been proposed that high resolution clustering, \emph{e.g.} obtained with Leiden or Louvain methods, can be refined in agglomerative way using machine learning techniques \cite{miao_2020}.

An alternative solution to community detection is the Stochastic Block Model, a generative model for graphs organized into communities \cite{holland_1983}. In this scenario, identification of cell groups requires the estimation of the proper parameters underlying the observed NN graph. According to the microcanonical formulation \cite{peixoto_2017}, the parameters are partitions and the matrix of edge counts between them. Under this model, nodes belonging to the same group have the same probability to be connected together. It is possible to include node degree among the model parameters \cite{karrer_2011}, to account for heterogeneity of degree distribution of real-world graphs. A Bayesian approach to infer parameters has been developed \cite{peixoto_2013} and implemented in the \texttt{graph-tool} python library (\href{https://graph-tool.skewed.de}{https:/\slash graph-tool.skewed.de}). There, a generative model of network $\boldsymbol A$ has a probability $P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b)$ where \textbf{$\boldsymbol\theta$} is the set of parameters and \textbf{\emph{$\boldsymbol b$}} is the set of partitions. The likelihood of the network being generated by a given partition can be measured by the posterior probability

@@

so that inference is performed by minimizing the information required to describe the data (Occam's razor); \texttt{graph-tool} is able to efficiently do this by a Markov Chain Monte Carlo approach \cite{peixoto_2014}. SBM itself may fail to identify small groups in large graphs, hence hierarchical formulation has been proposed \cite{peixoto_2014_h}. Under this model, communities are agglomerated at a higher level in a block multigraph, also modelled using SBM. This process is repeated recursively until a graph with a single block is reached, creating a Nested Stochastic Block Model (nSBM).

@@

\texttt{schist} is a convenient wrapper to the \texttt{graph-tool} python library, written in python and designed to be used with \texttt{scanpy}. The most prominent function is \texttt{schist.inference.nested\_model()} which takes a \texttt{AnnData} object as input and fits a nested Stochastic Block Model on the \emph{k}NN graph built with \texttt{scanpy} functions (\emph{e.g.} \texttt{scanpy.tools.neighbors()}). When launched with default parameters, \texttt{schist} fits a model which maximises the posterior probability of having a set of cell groups (or blocks) given a graph. \texttt{schist} then annotates cells in the data object with all the groups found at each level of a hierarchy. Given the large size of the NN graph in real-world experiments, it is possible that a single solution represents local minima of the fitting process. In addition, it is possible that multiple solutions are equally acceptable to represent the graph partitioning and a better description is given by the consensus over such solutions \cite{Peixoto_2021_consensus}. To overcome these issues, \texttt{schist} fits multiple instances in parallel and returns the inferred consensus model, alongside the marginal probabilities for each cell to belong to a specific group (\emph{cell marginals}). In addition to the default nested Stochastic Block Model, \texttt{schist} implements the possibility to fit a Bayesian model able to find statistically significant assortative modules \cite{Zhang_Peixoto_2020}, eliminating the need to set a resolution parameter as required in standard community detection by maximization of modularity.

@@

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{FigureTian.png}
\caption[]{Analysis of a mixture of 5 known cell lines. (A) UMAP embedding colored and labeled by cell line identity. H1975 cells (orange) show two distinct groups reflecting an internal heterogeneity. (B) UMAP embedding showing cells colored by level 1 of the hierarchy proposed by the nested Stochastic Block Model. (C) UMAP embedding showing cells colored according to the Leiden method at resolution $\gamma = 1$. (D) UMAP embedding colored by the classification made by SCCAF when partitions in (B) are used. (E) UMAP embedding colored by the classification made by SCCAF when partitions in (C) are used.}\label{FigureTian}
\end{figure}

@@

\subsection*{Hierarchy modeling complies with biological properties}

When grouping is performed by optimization of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, that is communities are consistently grouped at lower resolutions. Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas nSBM inherently encodes hierarchies by merging communities in a tree. The improper use of resolution parameter may lead to two types of errors: grouping of cells that are in fact distinct and creating an inconsistent hierarchy.

@@

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{Figure_Hemato_Lineage.png}
\caption[]{Analysis of hematopoietic differentiation. Each panel presents a low dimensional embedding of single cells next to a radial tree representation of the nSBM hierarchy. Cells are colored according to groupings at level 2 of the hierarchy, group 0 marks the most primitive population (A). In subsequent panels, cells are colored using a signature of Erythroid lineage (B), Monocytes (C) or Neutrophils (D).}\label{Figure_Hemato_Lineage}
\end{figure}

@@

We devised a simple metric, \emph{cell stability}, that is defined by the fraction of levels for which the marginal probability is higher than $1-1/B_i$. To do so, we only consider levels with at least two groups, hence excluding the root of the tree. We tested this metric on four datasets from \cite{mereu_2020} with different quality levels (iCELL8, MARS-seq, 10XV3 and Quartszeq-2) (Fig. \ref{Figure_Stability}). By taking a summary metric, \emph{e.g.} the mean $\overline{S}$ or the fraction of cells with $S>0.5$, we observed that it correlates with the data quality (Table \ref{table_stability}).

@@

The modelling approach we adopted allows the estimation of the information required to describe a graph given any partitioning scheme, not limited to the solution given by the model itself. Differences in entropy can be used to perform model selection, that is we can choose which model better describes the data. We sought to exploit this property to address the task of annotating cells according to a reference sample. To this end we analyzed datasets from \cite{mereu_2020}, which includes mixtures of human PBMC and HEK293T cells profiled with various technologies. We chose cells profiled with 10X V3 platform as reference dataset and performed annotation on cells profiled with Quartz-seq2 or MARS-seq. These are at the extremes of the capability to distinguish cell types, so they provide good benchmark configurations for this task.

After preprocessing raw data according to the parameters given in \cite{mereu_2020}, we integrated each dataset with 10XV3 into a unified representation using Harmony \cite{Korsunsky_2019}, and computed the \emph{k}NN graph. In each merged dataset, we retained cell type annotations for 10X cells, while we assigned a "Unknown" label to all cells derived from the other technology (\emph{i.e.} MARS-seq or Quartz-seq2). We then calculated the cell affinity matrix, that is we computed the difference in entropy that can be observed by assigning each cell to each annotation cluster, this being either one of the original cell types or "Unknown". Once the matrix has been computed, each cell from the query data is assigned to the group with the highest likelihood. The rationale behind this approach is that if cells belong to the same annotation group, then more information is required to describe the graph if they were annotated as different cell types; hence, cells from the query datasets should retain their "Unknown" label if and only if there is not enough evidence to associate them to another group. We compared the accuracy of the  outcome to \emph{k}NN classification, given by the closest entry in the \emph{k}NN graph, and to \texttt{ingest}, a tool included in \texttt{scanpy}  based on \emph{k}NN classification of UMAP embeddings. Analysis of a well defined dataset, such as Quartz-seq2, reveals that the three approaches are equally good in classifying unknown cells (Fig. \ref{Figure_Label_Transfer}, central column), with accuracies ranging from .870 to .927. When data are noisy, instead, \emph{k}NN-based methods show low accuracy and a tendency to assign the most represented cell group (HEK293T) to the unlabeled cells. This misannotation is particular evident for \texttt{ingest}, in which only CD4 T cells and HEK cells are transferred, resulting in the lowest accuracy (0.243). Conversely, \texttt{schist} is able to assign correct labels with higher accuracy (0.641). Moreover, \emph{k}NN methods assign a label to each cell, whereas \texttt{schist} does not relabel cells if there are no sufficient evidence (\emph{e.g.} the "Unknown" state is the most likely). Interestingly, we found that for the largest part of cells without assigned label, the second choice by affinity ranking was indeed the appropriate one (Fig. \ref{hist_label_second}).

@@

A possible strategy is based on Random Matrix Theory, as suggested by the authors of the SC3 package  \cite{kiselev_2017}, for which a suitable number of clusters, $\hat{k}$, is determined by the number of eigenvalues of the $\boldsymbol Z^\top\boldsymbol Z$ matrix (where $\boldsymbol Z$ is the normalized count matrix) significantly different at $p<.001$ from the appropriate Tracy-Widom distribution. According to this strategy, the optimal level $i^k$ is the one that minimizes the number of partitions and $\hat{k}$:

@@

An alternative strategy is to evaluate the behaviour of modularity at different hierarchy levels. While nSBM does not optimize the graph modularity $Q$, we observed that this tends to be maximal for the level better describing known cell populations, so the optimal level $i^Q$ is

@@

Where $Q_x$ is modularity at level $x$. We collected values arising from both the approaches for some datasets used in this work (Table \ref{table_optimal})

@@

Minimisation of the nSBM is a process that requires a large amount of computational resources. While the underlying \texttt{graph-tool} library is efficient in exploring the solution space using a multiflip MCMC sampling strategy, the number of required iterations before convergence is considerable and the running time scales linearly with the number of edges. Moreover, to collect a consensus partition, we minimize multiple models (default: 100) that need to be averaged. To give a reference, we report runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (Intel i7@2.8 GHz, 32 GB RAM). Compared to Leiden approach, nSBM requires at best $\sim6\times$ times more, and $\sim30\times$  at worst. A reasonably fast alternative to the nSBM is the Planted Partition Block Model (PPBM), for which we also report runtimes. The PPBM \cite{Zhang_Peixoto_2020} is able to find statistically significant assortative modules and eliminates the resolution parameter; differently from nSBM, PPBM is not hierarchic.

\begin{table}[h!]
\centering
 \begin{tabular}{|| l c c c c c ||}
%  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}
 \hline
 \textbf{Dataset} & \textbf{Cells} & \textbf{Edges} & \textbf{Leiden} & \textbf{PPBM} & \textbf{nSBM} \\ [0.5ex] 
 \hline\hline
 sc-mixology \cite{Tian_2019} & 860 & 9186 & 00:06 & 00:13 & 00:36 \\ 
 \hline
 Quartz-seq2 \cite{mereu_2020} & 1266 & 14603 & 00:10 & 00:19 & 00:45 \\
 \hline
 MARS-seq \cite{mereu_2020} & 1401 & 21756 & 00:20 & 00:34 & 02:14 \\
 \hline
 iCELL8 \cite{mereu_2020} & 1830 & 30636 & 00:23 & 00:40 & 03:02 \\
 \hline
 Chromium 10x \cite{mereu_2020} & 1523 & 21447 & 00:14 & 00:26 & 01:07 \\ 
% \hline
% Islets \cite{Segerstolpe_2016} & 2108 & 32023 & 00:20 & 00:42 & 03:13 \\
 \hline
 Hematopoiesis \cite{paul_2015} & 2730 & 15444 & 00:37 & 01:27 & 05:52\\ 
 \hline
 Mouse Cortex \cite{Zeisel_2015} & 3005 & 54460 & 00:59 & 00:53 & 07:32 \\
 \hline
 Endocrinogenesis \cite{Bastidas_Ponce_2019} & 3696 & 74670 & 01:15 & 01:29 & 10:56 \\
 \hline
 Baron Pancreas \cite{Baron_2016} & 8569 & 294480 & 03:51 & 07:35 & 1:33:40 \\
 \hline
 Airzani Liver \cite{Aizarani_2019} & 10368 & 354440 & 04:13 & 09:47 & 1:42:23 \\
 \hline
 Planaria \cite{plass_2018} & 21612 & 173667 & 05:41 & 13:52 & 1:20:40 \\
 \hline
\end{tabular}
\caption{Time required to run different partitioning strategies implemented in \texttt{schist} on various datasets. All approaches fit 100 models. Number of nodes and edges refer to the structure of the \emph{k}NN graph as built by \texttt{scanpy}. Times are expressed in mm:ss. }
\label{Table_runtime}
\end{table}

@@

The major drawback of adopting this strategy is the substantial increase of runtimes. As observed, model minimisation may be hundred times slower than the extremely fast Leiden approach. It should be noted that \texttt{schist} initializes multiple models that are treated by multiple concurrent processes. \texttt{graph-tool} itself supports CPU-level parallelization for some of its tasks. These optimizations are well suited for clustered computing infrastructure. Further development, possibly including GPU-level parallelization, is surely required to accomodate the large size of datasets that are being produced.

@@

Unless differently stated, all the analysis were produced using \texttt{scanpy} v1.7.1 \cite{wolf_2018} and \texttt{schist} v0.7.6 and the corresponding dependencies. All models were initialized 100 times, herein including Leiden partitioning for which we also calculated the consensus partition.

@@

Data and metadata for five cell mixture profiled by Chromium 10x were downloaded from the sc-mixology repository (\href{https://github.com/LuyiTian/sc_mixology}{https:/\slash github.com\slash LuyiTian\slash sc\_mixology}). Cells with less than 200 genes were excluded, as genes detected in less than 3 cells. Cells with less than 5\% of mitochondrial genes were retained for subsequent analysis. Data were normalised and log-transformed; number of genes and percentage of mitochondrial genes were regressed out. \emph{k}NN graph was built with default parameters (50 components and 15 nearest neighbors). Data were assessed by SCCAF using cell line annotation. Mean cross-validated accuracy was set as target for all the models.

\subsection*{Analysis of Visium H\&E data}
Data were retrieved using \texttt{squidpy.datasets.visium\_hne\_adata()} built-in function, without further processing. Leiden clustering was performed using \texttt{schist.inference.leiden()} function, allowing for 100 initializations, with resolutions $\gamma = 1$ and $\gamma = 0.5$.

@@

\subsection*{Processing of PBMC data from various platforms}
Count matrices were downloaded from GEO using the following accession numbers: GSE133535 (Chromium 10Xv3), GSE133543 (Quartz-seq2), GSE133542 (MARS-seq) and GSE133541 (iCELL8). Data were processed according to the methods in the original paper \cite{mereu_2020}. Briefly, cells with less than 10,000 total number of reads as well as the cells having less than 65\% of the reads mapped to their reference genome were discarded. Cells in the 95th percentile of the number of genes/cell and those having less than 25\% mitochondrial gene content were included in the downstream analyses. Genes that were expressed in less than five cells were removed. Data were normalized and log-transformed, highly variable genes were detected at minimal dispersion equal to 0.5. Neighbourhood graph was built using 30 principal components and 20 neighbours.

>>>>

A key step in the analysis of scRNA-seq data and, more in general, of single cell data, is the identification of cell populations, that is groups of cells sharing similar properties. Several approaches have been proposed to achieve this task, based on well established clustering techniques \cite{wang_2017, lin_2017}, consensus clustering \cite{huh_2020, kiselev_2017, Ranjan_2021} and deep learning \cite{li_2020}; many more have been recently reviewed \cite{krzak_2019, kiselev_2019} and benchmarked \cite{du_2018}. As the popularity of single cell analysis frameworks \texttt{Seurat} \cite{butler_2018} and \texttt{scanpy} \cite{wolf_2018} raised, methods based instead on graph partitioning became the \emph{de facto} standards. Such methods require the construction of a cell neighbourhood graph (\emph{e.g.} by \emph{k }Nearest Neighbours, \emph{k}NN, or shared Nearest Neighbours, \emph{s}NN). Encoding cell-to-cell similarities into graphs has practical advantages beyond clustering, as many algorithms for graph analysis can be applied and interpreted in a biological way. A notable example is the analysis of cell trajectories which can be derived from the analysis of Markov processes traversing the NN graph \cite{Setty_2019, Lange_2020}. In another context, computation of RNA moments in scRNA velocity is also based on the NN graph structure \cite{Bergen_2020}. Arguably, the biggest utility of NN structure is the possibility to identify cell groups by partitioning the graph into communities; this is typically achieved using the Louvain method \cite{blondel_2008}, a fast algorithm for optimisation of graph modularity. While fast, this method does not guarantee the identification of internally connected communities. To overcome its limits, a more recent approach, the Leiden algorithm \cite{traag_2019}, has been implemented and it has been quickly adopted in the analysis of single cell data, for example by \texttt{scanpy} \cite{wolf_2018} and \texttt{PhenoGraph} \cite{levine_2015}. In addition to Newman's modularity \cite{newman_2004}, other definitions currently used in single cell analysis make use of a resolution parameter \cite{traag_2011, reichardt_2006}. In lay terms, resolution works as a threshold on the density within communities: lowering the resolution results in less and sparser communities and \emph{vice versa}. Identification of an appropriate resolution has been recognised as a major issue \cite{lhnemann_2020}, also because it requires the definition of a mathematical property (clusters) over biological entities (the cell groups), with little formal description of the latter. In addition, the larger the dataset, the harder is to identify small cell groups, as a consequence of the well-known resolution limit \cite{fortunato_2007}. Moreover, it has been demonstrated that random networks can have modularity \cite{guimer_2004} and its optimisation is incapable of separating actual structure from those arising simply of statistical fluctuations of the null model. Lastly, it is a common error to assume that the resolution parameter reflects a hierarchical structure of the communities in the graph when, in general, this is not rigorously true. Additional solutions to cell group identification from NN graphs have been proposed, introducing resampling techniques \cite{baran_2019, Tang_Sackton_2020} or clique analysis \cite{xu_2015}. It has been proposed that high resolution clustering, \emph{e.g.} obtained with Leiden or Louvain methods, can be refined in agglomerative way using machine learning techniques \cite{miao_2020}.

An alternative solution to community detection is the Stochastic Block Model, a generative model for graphs organised into communities \cite{holland_1983}. In this scenario, identification of cell groups requires the estimation of the proper parameters underlying the observed NN graph. According to the microcanonical formulation \cite{peixoto_2017}, the parameters are partitions and the matrix of edge counts between them. Under this model, nodes belonging to the same group have the same probability to be connected together. It is possible to include node degree among the model parameters \cite{karrer_2011}, to account for heterogeneity of degree distribution of real-world graphs. A Bayesian approach to infer parameters has been developed \cite{peixoto_2013} and implemented in the \texttt{graph-tool} python library (\href{https://graph-tool.skewed.de}{https:/\slash graph-tool.skewed.de}). There, a generative model of network $\boldsymbol A$ has a probability $P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b)$ where \textbf{$\boldsymbol\theta$} is the set of parameters and \textbf{\emph{$\boldsymbol b$}} is the set of partitions. The likelihood of the network being generated by a given partition can be measured by the posterior probability

@@

so that inference is performed by minimising the information required to describe the data (Occam's razor); \texttt{graph-tool} is able to efficiently do this by a Markov Chain Monte Carlo approach \cite{peixoto_2014}. SBM itself may fail to identify small groups in large graphs, hence hierarchical formulation has been proposed \cite{peixoto_2014_h}. Under this model, communities are agglomerated at a higher level in a block multigraph, also modelled using SBM. This process is repeated recursively until a graph with a single block is reached, creating a Nested Stochastic Block Model (nSBM).

@@

\texttt{schist} is a convenient wrapper to the \texttt{graph-tool} python library, written in python and designed to be used with \texttt{scanpy}. The most prominent function is \texttt{schist.inference.nested\_model()} which takes a \texttt{AnnData} object as input and fits a nested Stochastic Block Model on the \emph{k}NN graph built with \texttt{scanpy} functions (\emph{e.g.} \texttt{scanpy.tools.neighbors()}). When launched with default parameters, \texttt{schist} fits a model which maximises the posterior probability of having a set of cell groups (or blocks) given a graph. \texttt{schist} then annotates cells in the data object with all the groups found at each level of a hierarchy. Given the large size of the NN graph in real-world experiments, it is possible that a single solution represents local minima of the fitting process. In addition, it is possible that multiple solutions are equally acceptable to represent the graph partitioning and a better description is given by the consensus over such solutions \cite{Peixoto_2021_consensus}. To overcome these issues, \texttt{schist} fits multiple instances in parallel and returns the inferred consensus model, alongside the marginal probabilities for each cell to belong to a specific group (\emph{cell marginals}). In addition to the default nested Stochastic Block Model, \texttt{schist} implements the possibility to fit a Bayesian model able to find statistically significant assortative modules \cite{Zhang_Peixoto_2020}, eliminating the need to set a resolution parameter as required in standard community detection by maximisation of modularity.

@@

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{FigureTian.png}
\caption[]{Analysis of a mixture of 5 known cell lines. (A) UMAP embedding coloured and labeled by cell line identity. H1975 cells (orange) show two distinct groups reflecting an internal heterogeneity. (B) UMAP embedding showing cells coloured by level 1 of the hierarchy proposed by the nested Stochastic Block Model. (C) UMAP embedding showing cells coloured according to the Leiden method at resolution $\gamma = 1$. (D) UMAP embedding coloured by the classification made by SCCAF when partitions in (B) are used. (E) UMAP embedding coloured by the classification made by SCCAF when partitions in (C) are used.}\label{FigureTian}
\end{figure}

@@

\subsection*{Hierarchy modelling complies with biological properties}

When grouping is performed by optimisation of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, that is communities are consistently grouped at lower resolutions. Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas nSBM inherently encodes hierarchies by merging communities in a tree. The improper use of resolution parameter may lead to two types of errors: grouping of cells that are in fact distinct and creating an inconsistent hierarchy.

@@

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{Figure_Hemato_Lineage.png}
\caption[]{Analysis of hematopoietic differentiation. Each panel presents a low dimensional embedding of single cells next to a radial tree representation of the nSBM hierarchy. Cells are coloured according to groupings at level 2 of the hierarchy, group 0 marks the most primitive population (A). In subsequent panels, cells are coloured using a signature of Erythroid lineage (B), Monocytes (C) or Neutrophils (D).}\label{Figure_Hemato_Lineage}
\end{figure}

@@

We devised a simple metric, \emph{cell stability}, that is defined by the fraction of levels for which the marginal probability is higher than $1-1/B_i$. To do so, we only consider levels with at least two groups, hence excluding the root of the tree. We tested this metric on four datasets from \cite{mereu_2020} with different quality levels (iCELL8, MARS-seq, 10XV3 and Quartsz-seq2) (Fig. \ref{Figure_Stability}). By taking a summary metric, \emph{e.g.} the mean $\overline{S}$ or the fraction of cells with $S>0.5$, we observed that it correlates with the data quality (Table \ref{table_stability}).

@@

The modelling approach we adopted allows the estimation of the information required to describe a graph given any partitioning scheme, not limited to the solution given by the model itself. Differences in entropy can be used to perform model selection, that is we can choose which model better describes the data. We sought to exploit this property to address the task of annotating cells according to a reference sample. To this end we analysed datasets from \cite{mereu_2020}, which includes mixtures of human PBMC and HEK293T cells profiled with various technologies. We chose cells profiled with 10X V3 platform as reference dataset and performed annotation on cells profiled with Quartz-seq2 or MARS-seq. These are at the extremes of the capability to distinguish cell types, so they provide good benchmark configurations for this task.

After preprocessing raw data according to the parameters given in \cite{mereu_2020}, we integrated each dataset with 10XV3 into a unified representation using Harmony \cite{Korsunsky_2019}, and computed the \emph{k}NN graph. In each merged dataset, we retained cell type annotations for 10X cells, while we assigned a "Unknown" label to all cells derived from the other technology (\emph{i.e.} MARS-seq or Quartz-seq2). We then calculated the cell affinity matrix, that is we computed the difference in entropy that can be observed by assigning each cell to each annotation cluster, this being either one of the original cell types or "Unknown". Once the matrix has been computed, each cell from the query data is assigned to the group with the highest likelihood. The rationale behind this approach is that if cells belong to the same annotation group, then more information is required to describe the graph if they were annotated as different cell types; hence, cells from the query datasets should retain their "Unknown" label if and only if there is not enough evidence to associate them to another group. We compared the accuracy of the  outcome to \emph{k}NN classification, given by the closest entry in the \emph{k}NN graph, and to \texttt{ingest}, a tool included in \texttt{scanpy}  based on \emph{k}NN classification of UMAP embeddings. Analysis of a well defined dataset, such as Quartz-seq2, reveals that the three approaches are equally good in classifying unknown cells (Fig. \ref{Figure_Label_Transfer}, central column), with accuracies ranging from .870 to .927. When data are noisy, instead, \emph{k}NN-based methods show low accuracy and a tendency to assign the most represented cell group (HEK293T) to the unlabelled cells. This misannotation is particular evident for \texttt{ingest}, in which only CD4 T cells and HEK cells are transferred, resulting in the lowest accuracy (0.243). Conversely, \texttt{schist} is able to assign correct labels with higher accuracy (0.641). Moreover, \emph{k}NN methods assign a label to each cell, whereas \texttt{schist} does not relabel cells if there are no sufficient evidence (\emph{e.g.} the "Unknown" state is the most likely). Interestingly, we found that for the largest part of cells without assigned label, the second choice by affinity ranking was indeed the appropriate one (Fig. \ref{hist_label_second}).

@@

A possible strategy is based on Random Matrix Theory, as suggested by the authors of the SC3 package  \cite{kiselev_2017}, for which a suitable number of clusters, $\hat{k}$, is determined by the number of eigenvalues of the $\boldsymbol Z^\top\boldsymbol Z$ matrix (where $\boldsymbol Z$ is the normalised count matrix) significantly different at $p<.001$ from the appropriate Tracy-Widom distribution. According to this strategy, the optimal level $i^k$ is the one that minimises the number of partitions and $\hat{k}$:

@@

An alternative strategy is to evaluate the behaviour of modularity at different hierarchy levels. While nSBM does not optimise the graph modularity $Q$, we observed that this tends to be maximal for the level better describing known cell populations, so the optimal level $i^Q$ is

@@

Where $Q_x$ is modularity at level $x$. We collected values arising from both the approaches for some datasets used in this work (Table \ref{table_optimal} and Fig. \ref{FigureOptimalLevel})

@@

Minimisation of the nSBM is a process that requires a large amount of computational resources. While the underlying \texttt{graph-tool} library is efficient in exploring the solution space using a multiflip MCMC sampling strategy, the number of required iterations before convergence is considerable and the running time scales linearly with the number of edges. Moreover, to collect a consensus partition, we minimise multiple models (default: 100) that need to be averaged. To give a reference, we report runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (Intel i7@2.8 GHz, 32 GB RAM). Compared to Leiden approach, nSBM requires at best $\sim6\times$ times more, and $\sim30\times$  at worst. A reasonably fast alternative to the nSBM is the Planted Partition Block Model (PPBM), for which we also report runtimes. The PPBM \cite{Zhang_Peixoto_2020} is able to find statistically significant assortative modules and eliminates the resolution parameter; differently from nSBM, PPBM is not hierarchic.

\begin{table}[h!]
\centering
 \begin{tabular}{|| l c c c c c ||}
%  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}
 \hline
 \textbf{Dataset} & \textbf{Cells} & \textbf{Edges} & \textbf{Leiden} & \textbf{PPBM} & \textbf{nSBM} \\ [0.5ex] 
 \hline\hline
 sc-mixology \cite{Tian_2019} & 860 & 9186 & 00:06 & 00:13 & 00:36 \\ 
 \hline
 Quartz-seq2 \cite{mereu_2020} & 1266 & 14603 & 00:10 & 00:19 & 00:45 \\
 \hline
 MARS-seq \cite{mereu_2020} & 1401 & 21756 & 00:20 & 00:34 & 02:14 \\
 \hline
 iCELL8 \cite{mereu_2020} & 1830 & 30636 & 00:23 & 00:40 & 03:02 \\
 \hline
 Chromium 10x \cite{mereu_2020} & 1523 & 21447 & 00:14 & 00:26 & 01:07 \\ 
% \hline
% Islets \cite{Segerstolpe_2016} & 2108 & 32023 & 00:20 & 00:42 & 03:13 \\
 \hline
 Hematopoiesis \cite{paul_2015} & 2730 & 15444 & 00:37 & 01:27 & 05:52\\ 
 \hline
 Mouse Cortex \cite{Zeisel_2015} & 3005 & 54460 & 00:59 & 00:53 & 07:32 \\
 \hline
 Endocrinogenesis \cite{Bastidas_Ponce_2019} & 3696 & 74670 & 01:15 & 01:29 & 10:56 \\
 \hline
 Baron Pancreas \cite{Baron_2016} & 8569 & 294480 & 03:51 & 07:35 & 1:33:40 \\
 \hline
 Airzani Liver \cite{Aizarani_2019} & 10368 & 354440 & 04:13 & 09:47 & 1:42:23 \\
 \hline
 Planaria \cite{plass_2018} & 21612 & 173667 & 05:41 & 13:52 & 1:20:40 \\
 \hline
\end{tabular}
\caption{Time required to run different partitioning strategies implemented in \texttt{schist} on various datasets. All approaches fit 100 models. Number of nodes and edges refer to the structure of the \emph{k}NN graph as built by \texttt{scanpy}. Times are expressed in MM:SS. }
\label{Table_runtime}
\end{table}

@@

The major drawback of adopting this strategy is the substantial increase of runtimes. As observed, model minimisation may be hundred times slower than the extremely fast Leiden approach. It should be noted that \texttt{schist} initialises multiple models that are treated by multiple concurrent processes. \texttt{graph-tool} itself supports CPU-level parallelisation for some of its tasks. These optimisations are well suited for clustered computing infrastructure. Further development, possibly including GPU-level parallelisation, is surely required to accomodate the large size of datasets that are being produced.

@@

Unless differently stated, all the analysis were produced using \texttt{scanpy} v1.7.1 \cite{wolf_2018} and \texttt{schist} v0.7.6 and the corresponding dependencies. All models were initialised 100 times, herein including Leiden partitioning for which we also calculated the consensus partition.

@@

Data and metadata for five cell mixture profiled by Chromium 10x were downloaded from the sc-mixology repository (\href{https://github.com/LuyiTian/sc_mixology}{https:/\slash github.com\slash LuyiTian\slash sc\_mixology}). Cells with less than 200 genes were excluded, as genes detected in less than 3 cells. Cells with less than 5\% of mitochondrial genes were retained for subsequent analysis. Data were normalised and log-transformed; number of genes and percentage of mitochondrial genes were regressed out. \emph{k}NN graph was built with default parameters (50 components and 15 nearest neighbours). Data were assessed by SCCAF using cell line annotation. Mean cross-validated accuracy was set as target for all the models.

\subsection*{Analysis of Visium H\&E data}
Data were retrieved using \texttt{squidpy.datasets.visium\_hne\_adata()} built-in function, without further processing. Leiden clustering was performed using \texttt{schist.inference.leiden()} function, allowing for 100 initialisations, with resolutions $\gamma = 1$ and $\gamma = 0.5$.

@@

\subsection*{Processing of PBMC data from various platforms}
Count matrices were downloaded from GEO using the following accession numbers: GSE133535 (Chromium 10Xv3), GSE133543 (Quartz-seq2), GSE133542 (MARS-seq) and GSE133541 (iCELL8). Data were processed according to the methods in the original paper \cite{mereu_2020}. Briefly, cells with less than 10,000 total number of reads as well as the cells having less than 65\% of the reads mapped to their reference genome were discarded. Cells in the 95th percentile of the number of genes/cell and those having less than 25\% mitochondrial gene content were included in the downstream analyses. Genes that were expressed in less than five cells were removed. Data were normalised and log-transformed, highly variable genes were detected at minimal dispersion equal to 0.5. Neighbourhood graph was built using 30 principal components and 20 neighbours.
